Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-08-04 Manuscript/proto-manuscript.tex
MS: updated and polished overall. Small updates to figures.

\begin{abstract}
  % insert abstract here
  \todo{Abstract directly from Joe's conference abstracts. To be rewritten.}
Classical molecular dynamics simulations give detailed information about membrane structure and dynamics. 
However, there is still a room for improvements in current force fields – it is known from the literature,
that the binding of ions, especially cations, to phopholipid membranes is overestimated in all classical models [1]. 
We suggest that the membrane-ion interactions can be corrected by including implicit electronic polarizability into the
lipid models through the electronic continuum correction (ECC) [2], which was already applied to monovalent and divalent
ions yielding models that feature correct ion pairing [3]. 
Using the electrometer concept [3, 4] and x-ray scattering form factors, our simulations point out that our hypothesis is
correct and ECC is indeed a missing important contribution in current classical lipid models. 
Moreover, the solid physical principles behind ECC are found not to hamper other relevant properties of a phospholipid bilayer. 
The new lipid model, "ECC-lipids", shows accurate binding affinity to sodium and calcium cations and headgroup order parameter response to bound charge. 
We also provide for the first time a realistic stochiometry of bound calcium cations to a POPC membrane, and their binding sites. 
This work will continue as an open collaboration project NMRlipids VI (http://nmrlipids.blogspot.fi).
\end{abstract}

@@

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
like in signal propagation in neurons and vesicle fusion.
Zwitterionic phosphocholine (PC) lipid bilayers have been widely used as model systems
to understand molecular level details of specific cation interactions with
cellular membranes by using experimental \cite{cevc90,tocanne90,binder02,pabst07,uhrikova08}
and theoretical methods \cite{bockmann03,bockmann04,Berkowitz12,??} \todo{We need to select references for these}.
While relative binding affinity of different ions is agreed to follow Hoffmeister
serie, the molecular details of binding and binding energetics are
not fully understood \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Non-invansive spectroscopic methods, like nucelar magnetic resosnance (NMR), scattering
and infrared scpectroscopy, give accurate information about ion binding
in lipid bilayers\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02}. 
Molecular level interpretation of the results, however, requires assumptions
about binding model and is often not fully conclusive \cite{??}.
On the other hand, the accuracy of the state of the art atomistic resolution lipid and ion models
have turned out insufficient for the detailed interpretation of cation binding details \cite{catte16}.

@@

In this work we show that the accuracy of Calcium binding
on zwitterionic PC lipid bilayer can be significantly improved by including electronic polarizability
by using effective continuum correction (ECC)~\cite{leontyev11} for polar region of zwitterionic lipid molecules
in classical MD simulation model.
This is essentially physically well justified version of partial charge scaling
implemented in early days of lipid and surfactant simulations \cite{jonsson86,egberts94}. 
The approach has been previously shown in to improve bulk performace of
ion models against neutron scattering data \cite{Jungwirth2015,kohagen14,kohagen16}.
The better bulk behaviour was not, however, sufficient to correct binding in lipid
bilayers \cite{catte16}. 
%motivation for its application to zwitterionic lipids like POPC.

%Classical molecular dynamics simulations have potential to solve the ion binding details
%in lipid bilayers and reveal its relevance to various biological
%problems \cite{bockmann03,bockmann04,??}. However, the available
%molecular dynamics simulation models have a strong tendency to overestimate
%cation binding on zwitterionic bilayers and none of them can reproduce the
%experimental data with the accuracy required for the interpretation of
%experiments \cite{catte16}.

The proposed MD simulation model with ECC-corrected lipids and ions
reproduce the experimental NMR data of Calcium binding in PC lipid bilayers
and mainly supports the proposed ternary complex model (2 Ca2+ bind in
on POPC) \cite{akutsu81,altenbach84}. However, non-negligible contribution
of one Ca2+ cross bridging three POPC molecules is also observed.
The overestimated specific cation binding observed in current lipid models \cite{catte16}
may lead to articifially positively charged memranes and significant artefacts in
MD simualtions. Thus, the proposed approach to improve the accuracy of cation
binding in current MD models is highly useful for future MD simulations
with physiological salt conditions.

%re is a vast demand for classical molecular dynamics simulation
%model which correctly reproduces cation binding in lipid bilayers.

@@

\subsection{Electronic continuum correction for lipid bilayers}
The lack of electronic polarizability from the classical MD simulation
models has been considered higly relevant issue since the early days of
lipid bilayer simulations and pioneering simulation studies scaled
the partial charges by half to effectively include electronic
polarizability \cite{jonsson86,egberts94}. Also approaches that
explicitly include electronic polarizability has been introduced, but
this has turned out to be practically complicated \cite{chowdhary13}.
In this work we effectively include electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. The approach is more sophisticated version of
early studies with scaling to half was used \cite{jonsson86,egberts94},
but significantly more simple than the explicit inclusion of electronic
polarizability \cite{chowdhary13}.

%is a potential source of artefact, which could lead to overbinding
%of cations. 
%Almost all lipid bilayer simulations reported in the literature 
%are done with force fields that represent interactions between atoms
%with pairwise additive empirical potential and exclude the electronic 
%polarizability \cite{SOME_review, FF_papers}.

@@

The main goal of this work is to apply ECC correction to accurately 
describe the lipid headgroup response to Na$^+$ and Ca$^{2+}$ concentrations
in MD simulation models when compared with NMR data \cite{catte16}. 
The Lipid14~\cite{dickson14} force field parameters were used as a starting point,
because their response to bound ions was most realistic against NMR data
(see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and headgroup structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.

While the scaling factor of $f_q = 0.75$ for ions in water is well justified and shown
to improve model performance against scattering data \cite{kohagen14,kohagen16,??},
it is not clear if the same factor should be used for lipid headgroups \todo{, because ??}.
Here we estimated the correct scaling factor for Lipid14 model by using
so called implicitly polarized charges (IPolQ) \cite{ipolq2013},
which are the average of partial charges given by RESP~\cite{RESP_paper} calculations in
vacuum and in implicit solvent.
%In order to implicitly capture solvent induced polarization 
%in models with fixed charges, 
%we shall use the average of partial charges 
%from both environments (i.e. vaccum and solvent),
%so called implicitly polarized charges (IPolQ) \cite{ipolq2013}. 
The partial charges in original Lipid14 parametrization were derived by fitting the electrostatic 
potential to its model quantum chemistry representation (RESP\cite{RESP_paper}) in vacuum,
while charges from calculations in an implicit solvent vary most for the polar moieties \cite{maciejewski14}. 
Here we estimate the appropriate scaling factor for lipid hedgroups and glycerol backbone
by comparing IPolQ charges (average of RESP calculation in vacuum and in implicit solvent) to
the RESP results in vacuum. By using the results reported in \cite{maciejewski14} we conclude
that the IPolQ charges can be achived from vacuum calculations by scaling roughly with the
factor of $f_q=0.8$, which is the used to derive ECC-corrected Lipid14 model.
\todo{SAMULI: please check that this is correctly descrived.}
%By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
%we represent the IPolQ charges in the electronic continuum correction of the Lipid14 model
%by increasing the scaling factor $f_q$ to $0.8$. 
\footnote{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  I use 0.80 as an approximate round value. The use of 0.78---.79 might be more appropriate, though.}

Here we apply ECC correction with the proposed scaling factor only to the headgroup,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
highly optimized in Lipid14 and other lipids models and
give generally a good description for hydrophobic part of lipid bilayers
in various conditions \cite{ollila16},
in contrast to glycerol backbone and headgroup regions which require
some improvement in all available lipid models \cite{botan15}. 
%To minimize the detuning of the highly optimized hydrocarbon
%parameters, 
%Because the biologically relevant cations like \ce{Na^+} and \ce{Ca^{2+}}
%are found to interact only with the phospholipid head group,
%and the accuracy of alkyl sidechains in Lipid14 model are satisfactory, 
%we exclude alkyl tails from our modifications. 
The scaling of partial charges in the polar region was found to decrease
area per molecule to ??\todo{find the value}, which is significantly smaller than the
experimental value (\cite{}) and the original Lipid14 values (\cite{})\todo{Add the values}.  
The decrease of area was found to arise from lower hydration of the lipid headgroup region,
%\todo{SAMULI: Did you analyze the effect of this to hydration or did we only speculate?
%JOE: I found a systematic decrease of water density in the heagroup region even if I kept APL constant.}
which can be explained by the increased solvation free energy due to the lower polarity
of molecules with scaled charges. The solvation free energy can be decreased
and hydration increased by reducing the radius of beads by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms.
%This allows water molecules to approach closer to lipid atoms and
%have stonger electrostatic interactions with them.
%\todo{SAMULI: This effect may have an official name. In that case we should mention it.}
Here we reduce $\sigma$ with a scaling factor of $f_\sigma = 0.89$ for the same atoms for which
charges was scaled. This increased the area per molecule back
to a level close to experimental and original Lipid14 values. 
%\todo{Is there some justification for the below sentence?} 
%In line with the electronic continnum correction,
%the scaling factor for sigma parameters, $f_\sigma$,
%has to lay between $f_q$ and 1.

\subsection{Comparison of ion binding affinity to experiments by using the electrometer concept}
Ion binding was compared between experiments and simulations by 
using lipid headgroup order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid headgroup
(see Fig. \ref{??} 
\todo{Figure with chemical structure and labeling to be added}) 
are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}.
More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid headgroup dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
Order parameters for C-H bonds can be accurately measured for each
C-H bond of lipids by using H$^2$ NMR or $^{13}$C NMR techniques and
calculated from MD simulations directly from definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

The change of the headgroup order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Combination of atomic absorption spectra, $^2$H NMR experiments
and information about order parameter signs gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.

Recent work published by the NMRlipids project showed that the concept works
qualitatively also in simulations and can be used to compare ion binding
affinitity between simulations and experiments~\cite{catte16}. However,
it also turned out that also the sensitivity of the order parameters response
to bound charge has to be quantified more accurately for quantitative comparison
of binding affinity. In this work we first we first quantify the response of
headgroup order parameters against experimental data from cationinc surfactants
embedded in PC bilayer~\cite{scherer89}. Essentially all surfactants locate in
lipid bilayers in these experiments, thus the amount of bound charge is known
and can be plotted against order parameter change.

@@

\subsection{Comparison of lipid bilayer structure to experimental data}
Lipid bilayer structure without ions was validated against NMR experiments
by order parameters for C-H bonds and x-ray scattering experiments by using
form factors. The former validates the structures sampled by individual
lipid molecules in simulations with segmental resolution, while latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}. Form factor was calculated from simulations
as 
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering lenght,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning the membrane with thickness $D$.

@@

\subsubsection{Simulations with aqueous ions}
The simulated systems consisted of 1-Palmitoyl-2-oleoylphosphatidylcholine
(POPC) bilayer and an aqueous salt solutions of varying concentrations.
Water molecules were described by OPC3 model \cite{Izadi16}.
In order to test transferability of the newly developed ECC-lipids model, 
we also performed several additional simulations with water models 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987} and TIP4p/2005 \cite{Abascal2005}
presented in Supporting Information (SI). 
We used ECC-ions model for ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion model by Dang \cite{smith94,chang1999,dang2006}. 
\todo{Table about simulations details and some related text is deleted in commit
  https://github.com/ohsOllila/NMRlipids\_VI-NewIonModel/commit/6526079062ca322e90f8ea0654597394b023041e
  Maybe this is done accidentally, or is there some other plan with this?
}

@@

\subsection{Lipid bilayers without ions}
The scattering form factors, NMR order parameters and area per lipid calculated  
using the ECC-lipids model for POPC are compared to experiments and original
Lipid14 results in Fig. \ref{simVSexpNOions} and in Table~\ref{tab:apls}. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.
The structural quality is comparable to the state of art lipid models available in literature \cite{ollila16},
thus we conclude that the ECC-lipids model reproduces the lipid bilayer structure
in liquid disordered phase with similar accuracy to other available models.

>>>>

\begin{abstract}
  % insert abstract here
  \todo{Abstract directly from Joe's conference abstracts. To be rewritten.}
Classical molecular dynamics simulations give detailed information about membrane structure and dynamics. 
However, there is still a room for improvements in current force fields – it is known from the literature,
that the binding of ions, especially cations, to phopholipid membranes is overestimated in all classical models [1]. 
We suggest that the membrane-ion interactions can be corrected by including implicit electronic polarizability into the
lipid models through the electronic continuum correction (ECC) [2], which was already applied to monovalent and divalent
ions yielding models that feature correct ion pairing [3]. 
Using the electrometer concept [3, 4] and x-ray scattering form factors, our simulations point out that our hypothesis is
correct and ECC is indeed a missing important contribution in current classical lipid models. 
Moreover, the solid physical principles behind ECC are found not to hamper other relevant properties of a phospholipid bilayer. 
The new lipid model, "ECC-lipids", shows accurate binding affinity to sodium and calcium cations and head group order parameter response to bound charge. 
We also provide for the first time a realistic stochiometry of bound calcium cations to a POPC membrane, and their binding sites. 
This work will continue as an open collaboration project NMRlipids VI (http://nmrlipids.blogspot.fi).
\end{abstract}

@@

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
like in signal propagation in neurons and vesicle fusion.
\todo{JOE: list more introductory examples, where it has interesting effects.}
Zwitterionic phosphocholine (PC) lipid bilayers have been widely used as model systems
to understand molecular level details of specific cation interactions with
cellular membranes by using experimental \cite{cevc90,tocanne90,binder02,pabst07,uhrikova08}
and theoretical methods \cite{bockmann03,bockmann04,Berkowitz12,??} \todo{We need to select references for these}.
While relative binding affinity of different ions is agreed to follow Hoffmeister
series, the molecular details of binding and binding energetics are
not fully understood \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Non-invansive spectroscopic methods, like nucelar magnetic resosnance (NMR), scattering
and infrared scpectroscopy, give accurate information about ion binding
in lipid bilayers\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02}. 
Molecular level interpretation of the results, however, requires assumptions
about binding model and is often not fully conclusive \cite{??}.
On the other hand, the accuracy of the state of the art atomistic resolution lipid and ion models
have turned out insufficient for the detailed interpretation of cation binding details \cite{catte16}.
The overestimated specific cation binding observed in current lipid models \cite{catte16}
may lead to articifially positively charged memranes and significant artefacts in
MD simualtions (e.g.~\ce{Ca^2+} induced charge inversion of bacterial membranes~\cite{Luan2016}).

@@

In this work we show that the accuracy of ion binding, especially Calcium,
in zwitterionic PC lipid bilayer can be significantly improved 
for classical MD simulation models of zwitterionic lipids
by including electronic polarizability
in the polar region of lipid molecules
using the electronic continuum correction (ECC)~\cite{leontyev11}. 
This approach has been previously shown to improve performace of
ion models against neutron scattering data in concentrated solutions \cite{Jungwirth2015,kohagen14,kohagen16}
and ab-initio simulations~\cite{Jungwirth2017}. 
The realistic structure of ions in a bulk solvent was not, however, sufficient to correct binding in lipid
bilayers~\cite{catte16}. 
To fix it, we propose
a classical fixed-charge MD simulation model of 
1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) lipid 
that accounts for electronic polarizability by using ECC. 
We will validate such a model by comparing its lipid bilayer structue
to x-ray scattering data and its head group structure to NMR experiments. 
We will then quantify the accuracy of head group order parameter response to bound charge
and compare it to the experimental NMR data of Sodium and Calcium binding in PC lipid bilayers. 
In addition, we will quantify relative binding affinities of different moieties in POPC to such cations. 
This will be discussed in terms of binding modes, stoichiometry and binding isotherms and their models used in experiments.
The proposed POPC model, ECC-lipids, and the approach to obtain it
is, hence, highly useful for future MD simulations with physiological salt conditions, 
and for effects where membrane electrostatics play a major role.

@@

\subsection{Electronic continuum correction for lipid bilayers}
The lack of electronic polarizability in the classical MD simulation
models has been considered higly relevant issue since the early days of
lipid bilayer simulations. Some pioneering simulation studies scaled
the partial charges of atoms by one half to effectively include 
polarizability \cite{jonsson86,egberts94}. Also, approaches that
explicitly include polarizability has been introduced, but
this has turned out to be practically complicated \cite{chowdhary13}.
In this work we effectively include electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. 
Technically, it is a similar approach to the phenomenological charge-scaling as applied in 
the early studies where a scaling factor one half was used \cite{jonsson86,egberts94}. 
In contrast to those studies, however, the concept of ECC is 
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11},
but significantly more simple than the explicit inclusion of electronic
polarizability \cite{chowdhary13} both for derivation and application.

@@

The main goal of this work is to apply ECC correction to accurately 
describe the lipid head group response to Na$^+$ and Ca$^{2+}$ concentrations
in MD simulation models when compared with NMR data \cite{catte16}. 
The Lipid14~\cite{dickson14} force field parameters were used as a starting point,
because their response to bound ions was most realistic against NMR data
(see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.

While the scaling factor of $f_q = 0.75$ for ions in water is well justified and shown
to improve model performance against scattering data \cite{kohagen14,kohagen16, Pluharova2014},
it is not clear if the same factor should be used for lipid head groups. 
Unlike the total charge of cations and anions, atomic partial charges used in
classical MD simulations are not experimental observables and 
can in principle adopt arbitrary values. 
Here we estimate the correct scaling factor for Lipid14 model by using
so called ``implicitly polarized charges'' (IPolQ) \cite{ipolq2013},
which are the average of partial charges given by 
restrained electrostatic potential fitting (RESP~\cite{RESP_paper}) 
in vacuum and in solvent.
Using the partial atomic charges obtained from RESP fitting in vacuum and implicit solvent reported in \cite{maciejewski14} and
considering that the partial charges in original Lipid14 parametrization were derived using RESP~\cite{RESP_paper} in vacuum,
we conclude that the charges in Lipid14 model shall be scaled roughly by the factor $f_q=0.8$. 
%By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
%we represent the IPolQ charges in the electronic continuum correction of the Lipid14 model
%by increasing the scaling factor $f_q$ to $0.8$. 
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

Here we apply the ECC correction with the proposed scaling factor only to the head group,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
already highly optimized in most lipids models including Lipid14 and
give generally a good description for hydrophobic part of lipid bilayers
in various conditions \cite{ollila16}.
In contrast, all available lipid models require
some improvements in glycerol backbone and head group regions \cite{botan15}. 
%To minimize the detuning of the highly optimized hydrocarbon
%parameters, 
%Because the biologically relevant cations like \ce{Na^+} and \ce{Ca^{2+}}
%are found to interact only with the phospholipid head group,
%and the accuracy of alkyl sidechains in Lipid14 model are satisfactory, 
%we exclude alkyl tails from our modifications. 
The scaling of partial charges in the polar region was found to decrease
area per molecule to 60~\AA$^2$, which is significantly smaller than the
experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment}) and the original Lipid14 values $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}..  
The decrease of area was found to arise from a lower hydration of the lipid head group region,
%\todo{SAMULI: Did you analyze the effect of this to hydration or did we only speculate?
%JOE: I found a systematic decrease of water density in the heagroup region even if I kept APL constant.}
which can be explained by the decreased solvation free energy due to the lower polarity
of molecules with scaled charges. 
The solvation free energy and hydration is sensitive to Van der Waals interaction parameters,
hence, they can be increased by reducing the radius of the selected atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential. 
%This allows water molecules to approach closer to lipid atoms and
%have stonger electrostatic interactions with them.
%\todo{SAMULI: This effect may have an official name. In that case we should mention it.}
Here we reduce $\sigma$ with a scaling factor of $f_\sigma = 0.89$ for the same atoms for which
charges were scaled. This has lead to an increase of the area per molecule back
close to the experimental value (see Table~\ref{tab:apls}). 
%\todo{Is there some justification for the below sentence?} 
%In line with the electronic continnum correction,
%the scaling factor for sigma parameters, $f_\sigma$,
%has to lay between $f_q$ and 1.

\subsection{Comparison of ion binding affinity to experiments by using the electrometer concept}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid head group
(see Fig. \ref{??} 
\todo{Figure with chemical structure and labeling to be added}) 
are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}.
More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid head group dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
Order parameters for C-H bonds can be accurately measured for each
C-H bond of lipids by using H$^2$ NMR or $^{13}$C NMR techniques and
calculated from MD simulations directly from definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

The change of the head group order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Combination of atomic absorption spectra, $^2$H NMR experiments
and information about order parameter signs gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.

Recent work published by the NMRlipids project showed that the concept works
qualitatively also in simulations and can be used to compare ion binding
affinitity between simulations and experiments~\cite{catte16}. However,
it also turned out that the sensitivity of the order parameters response
to the bound charge has to be quantified more accurately for quantitative comparison
of binding affinity. Hence, we first quantify the response of
head group order parameters against experimental data from cationinc surfactants
embedded in PC bilayer~\cite{scherer89}. Essentially all surfactants locate in
lipid bilayers in these experiments, thus the amount of the bound charge is known
and can be plotted against order parameter change.
Secondly we evaluate the response of the head group order parameters to varying salt concentrations in solution
and estimate the amount of membrane bound charge.

@@

\subsection{Comparison of lipid bilayer structure to experimental data}

\begin{figure}[tbp]
  \centering
  \includegraphics[height=8.5cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \includegraphics[width=9.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.png}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.}
  \todo{Add acyl chain order parameters, POPC chemical structure} \\
  \todo{We should run a simulation for the system withtout ions with the
    same (or close) temperature as the experimental data. The NMR data is at 300K
    so that would be good. I think that this would be also close enough for the scattering data at 303K.
  }
\end{figure}

Lipid bilayer structure without ions was validated against NMR experiments
by order parameters for C-H bonds and x-ray scattering experiments by using
form factors. The former validates the structures sampled by individual
lipid molecules in simulations with segmental resolution, while the latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}. Form factor was calculated from simulations
as 
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering lenght,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning the membrane with thickness $D$.

The scattering form factors, NMR order parameters and area per lipid calculated  
using the ECC-lipids model for POPC are compared to experiments and original
Lipid14 results in Fig. \ref{simVSexpNOions} and in Table~\ref{tab:apls}. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.
The structural quality is comparable to the state of art lipid models available in literature \cite{ollila16},
thus we conclude that the ECC-lipids model reproduces the lipid bilayer structure
in liquid disordered phase with similar accuracy to other available models.

@@

\subsubsection{Simulations with aqueous ions}
The simulated systems consisted of 1-Palmitoyl-2-oleoylphosphatidylcholine
(POPC) bilayer and an aqueous salt solutions of varying concentrations.
Water molecules were described by OPC3 model \cite{Izadi16}.
In order to test transferability of the newly developed ECC-lipids model, 
we also performed several additional simulations with water models 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987} and TIP4p/2005 \cite{Abascal2005}
presented in Supporting Information (SI). 
We used ECC-ions model for ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion model by Dang \cite{smith94,chang1999,dang2006}. 
Classical molecular dynamics simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}, 
and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}. 
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

\begin{table}[btp]
  \caption{Simulation parameters}
  \label{tbl:mdpar}
  \begin{tabular}{ll}
    simulation property & parameter   \\
    \hline
    time-step           & 2~fs         \\
    equilibration time  & 100~ns  \\
    simulation time     & 200~ns  \\
    temperature         & 313~K       \\
    thermostat          & v-rescale  \cite{bussi07}   \\
    barostat            & Parrinello-Rahman, semi-isotropic \cite{parrinello81} \\
    long-range electrostatics & PME  \cite{darden93}  \\
    cut-off scheme      & Verlet \cite{Pall13}      \\
    Coulomb and VdW cut-off & 1.0~nm \\
    constraints         & LINCS, only Hydrogen atoms \cite{hess97} \\
    constraints for water & SETTLE  \cite{miyamoto92} \\
    \hline
  \end{tabular}
\end{table}

@@

\subsection{POPC membrane and electrometer concept with cationic surfactant}

\begin{figure*}[tbp]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OrdParChanges_NaCl_CaCl2_surf.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Headgroup order parameter changes as a function of NaCl, CaCl$_2$ concentration and
    cationic surfactant (dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) mole fraction
    from simulations and experiments (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, surfactant \cite{scherer89}).
  }
\end{figure*}

Headgroup order parameter response to the bound charge was evaluated against
experimental data measured with the cationic surfactant
(dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) \cite{scherer89}.
The exact amount of bound charge in the membrane is known in these systems, because
practically all cationic surfactant molecules are embedded in the lipid bilayer due to
their amphiphilic nature. Thus, the changes of head group order parameter as a function of
mole fraction of cationic surfactants (Fig. \ref{OrderParameterCHANGESnewMODELS}) gives
also the order parameter changes as a function of the amount of bound cations.
\todo{Split the OP-changes figure in two: cationic surfactant -and- binding of cations.
      The separate figure with the cationic surfactant shall be additionally accompanied
      with a plot of Head group tilt (PN-vector?) change/response. }
